---
title: "Analysis of Various Factors Causing Deaths in Aircraft Accidents"
author: "By Madhur Bansal, Sambit Das, Sonali Saren"
date : Group 17
output: pdf_document
geometry: "left = 2.5cm, right = 2.5cm, top = 2cm, bottom = 2cm"
fontsize: 14pt
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

\pagenumbering{gobble}

```{=tex}
\begin{centering}
\vspace{3cm}

Under the guidance of Dr. Dootika Vats

\end{centering}
```
\newpage

```{=tex}
\pagenumbering{arabic}
\begin{center} \Large \textbf{Acknowledgement}\end{center}
```
We group-17 would like to express our profound gratitude towards *Dr.
Dootika Vats*, our academic and project instructor for MTH208A(Data
Science Lab), for her guidance and constant supervision throughout the
process and providing creative ideas and necessary information regarding
the project which led to the completion of this project. It has been a
great learning experience and has also provided us with a practical
insight of the theoretical knowledge gathered during the course lecture.

\newpage

```{=tex}
\begin{center} \Large \textbf{Abstract}\end{center}
```
This project deals with analysing data on aircraft accidents across
various locations over a period of 1938 to 2022. The motive of this
project is to understand the relation between the numbers of casualties
for different parameters of aircraft accidents like type of the
accident, phase of the accident, location of the accident, etc. We have
also tried to figure out from the data whether we can filter out
information and conclude anything on the progress of the aviation industry , by studying the trend of the total number of casualties with time and how they vary for various aircraft manufacturers.

\newpage


```{=tex}
\begin{center} \Large \textbf{Introduction}\end{center}
```
There are various types of vehicular accidents that can occur; aircraft
accidents being a very serious one of them. While it is unfortunate that
these accidents occur, air travel only gets more reliable as the
previous accidents result in more crucial safety improvements. There are
various reasons for such accidents. Defects in aircrafts can cause
equipment failures leading to malfunctioning of some parts, thus causing
accidents. Other reasons like human errors caused by mechanics, ground
workers and sometimes crew members also lead to mishaps leading to
casualties. Analyzing data on aircraft accidents can give us valuable
insights and important information on various factors influencing the
number of such casualties.

\newpage

```{=tex}
\begin{center} \Large \textbf{Data Description}\end{center}
```
**Data Collection**

The data was collected from Wikipedia using techniques of web scraping
in R programming language. Libraries like tidyverse and rvest and
functions like html_elements and html_table were used to scrape the data
from the webpage.

**Description**

The final workable data we obtained after cleaning contains 544
observations ranging over a period from 1938 to 2022 and providing
information about the following:

1.  Number of casualties further divided into four parts:
    -   Total number of deaths
    -   Number of deaths of the crew members
    -   Number of deaths of the passengers
    -   Number of deaths of the ground workers
2.  The type of aircraft accident
3.  The name of the manufacturers of the aircrafts met with the
    accidents
4.  The location of the accident which gives us information about the
    country in which the accident took place.
5.  Phase of the accident
6.  Year of the accident

There are six types of aircraft accidents in our data. They are
explained as follows:

a)  COM: It is the abbreviated form of commercial type of accidents. It
    takes into account all the accidents involving a commercial flight.
b)  EXG: It refers to the accidents where the aircraft is attacked
    externally using ground based weapons like ground to air missiles.
c)  EXG: It refers to the accidents where the aircraft is harmed using
    attacks originating in the sky. For example intentional downing by a
    military aircraft will fall under this category.
d)  INB: It refers to accidents resulting from internal attacks
    involving a pre-planned bomb, without hijacking.
e)  INH: It refers to the accidents resulting from internal attacks to
    commandeer an aircraft. Weapons used for this purpose including
    explosives is coded in this category.
f)  MIL: It refers to the accidents involving military aircrafts.

There are nine phases of accidents considered in our data. They are
explained as follows:

a)  STD: This is the standing phase of the flight that is prior to
    pushback or after gate arrival or when parked.
b)  TXI: This is the taxi phase which is prior to take-off. In this
    phase the flight moves under its own power.
c)  TOF: This is the take-off phase starting from the initiation of
    take-off power through to 10m altitude.
d)  ICL: ICL refers to the initial climb phase which marks the end of
    the take-off phase. During this period, the aircraft climbs to a
    pre-determined cruising altitude (approximately 300m)
e)  ENR: This is the en-route phase which starts after the initial climb
    phase and continues till the initiation of the approach phase.
f)  MNV: MNV stands for maneuvering. It is the act of flying an aircraft
    close to the ground involving steep turns and aerobatics.
g)  APR: This is the approach phase where the aircraft gradually slows
    down and prepares for landing.
h)  LDG: This is the landing phase of flight where the aircraft lands
    and exits the runway.
i)  UNK: This is the abbreviated form for those accidents for which the
    phase of the flight could not be determined.

\newpage

```{=tex}
\begin{center} \Large \textbf{Key Questions}\end{center}
```
We aim to provide answer to the following questions from our analysis of
the data:

1.  How is the number of deaths related to different parameters like
    type of accidents, phase of accidents and location of accidents?
2.  What is the trend of the number of deaths with time?
3.  Which airline company is the safest to travel with?
4.  Which manufacturer have made the most improvement in terms of safety
    over the period of time?

\newpage

```{=tex}
\begin{center} \Large \textbf{Insightful Visualizations}\end{center}
```
Studying and analyzing data is tough without plotting since data
visualization help us to notice every difference with respect to every
variable and also helps us to study trends, quantifying relationships,
or displaying proportions.Plots which are self explanatory are the
**best plots**.

We started analyzing the relationship of the number of deaths with
different factors by plotting it against them.

The **first plot** gives us an insight of the relationship between the
number of deaths with the different types of aircraft accidents.

```{r, echo=FALSE}
# Death vs Type plot
load("D:/SAMBIT/M.Sc/Sem 1/Data Science/R_IITK/PROJECT/Airplane_crashes.Rdata")
library(ggplot2)
g <- ggplot(final_df, aes(Type, `Total deaths`, fill = Type))
g + geom_bar(stat="identity", width = 0.5) +
  labs(title="Aircraft accident type vs Total deaths") +
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
  theme_minimal() +
  labs(x = "Accident Type") +
  scale_fill_discrete(name = "Accident Type", labels = c("COM-Commercial Accidents", "EXG-Attacked using weapons", "EXS-Attacked by aircrafts", "INB-Bombing", "INH-Hijacking", "MIL-Military accidents"))

```


From the graph it is clear that commercial type of accidents contribute
to the most number of deaths. One probable reason for this might be the
fact that commercial flights are the most flown. Also it carries many
passengers at a time, so an accident in a commercial flight is more
likely to result in more number of deaths.We can see **Commercial**
accidents caused nearly 45000 deaths, followed by **Military** accidents
which caused about 5000 deaths, then by **Hijacking**, **Bombimg** and
others. Hijacking and bombing kind of incidents are rare, hence
contributing to a lower number of deaths.

\newpage

The **second plot** gives us an insight on how the total number of
deaths varies with the different phases of accidents.

```{r echo=FALSE}
g <- ggplot(final_df, aes(Phase, `Total deaths`, fill = Phase))
g + geom_bar(stat="identity", width = 0.5) +
  labs(title="Phase of accident vs Total Deaths") +
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
  theme_minimal() +
  labs(x = "Phase of Accident") +
  scale_fill_discrete(name = "Accident Type", labels = c("Approach", "En Route", "Initial Climb", "Landing", "Maneuvering", "Standing", "Take off", "Taxi", "Unknown"))

```


From the graph it can be seen that accidents during the en-route phase
have resulted in the maximum number of deaths followed by other phases
like approach, initial climb and others. During the en-route phase, the
flight is in mid air. Hence an accident during this point of no return
is likely to be more fatal than accidents in other phases. Also we see
that the number of deaths for the standing phase is the least which is
obvious since during the standing phase the aircraft is not in motion,
hence the chance of an accident is very low during this phase.

\newpage

To study how **aircraft accidents** have caused deaths across different
**locations** of the world, we have plotted a map for the total number of
deaths corresponding to the different countries in our data.

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide',fig.keep='all'}
library("tidyverse") # load dplyr, ggplot2, stringr, etc.
library("countrycode") # get ISO code from country names
library("rworldmap")

data_with_iso <- final_df %>%
  mutate(Iso3 = countrycode::countrycode(
    sourcevar = Location, 
    origin = "country.name", 
    destination = "iso3c")
  )

#data_with_iso

data_country <- data_with_iso %>% group_by(Iso3) %>% summarise(sum(`Total deaths`), 
                                                       sum(`Crew deaths`),
                                                       sum(`Passenger deaths`),
                                                       sum(`Ground deaths`))
colnames(data_country) <- c("Iso3", "Total deaths","Crew deaths", "Passenger deaths", "Ground deaths")


sPDF <- joinCountryData2Map( data_country,
                             joinCode = "ISO3",
                             nameJoinColumn = "Iso3" )

#creating a user defined colour palette
op <- palette(c("lightgreen","yellow","orange","red"))
#find quartile breaks
cutVector <- c(50, 400, 800, 1800, 9000)
#classify the data to a factor
sPDF@data[["Deaths"]] <- cut(sPDF@data[["Total deaths"]]
                                        , cutVector
                                        , include.lowest=TRUE )
levels(sPDF@data[["Deaths"]]) <- c("low", "med", "high", "vhigh")
#mapping
mapCountryData( sPDF , nameColumnToPlot="Deaths"
                    , catMethod= "categorical"
                    , mapTitle= "Deaths vs Country"
                    , colourPalette= "palette"
                    , oceanCol= "lightblue"
                    , missingCountryCol = "white")
```


Here the red color denotes very high number of deaths, orange denotes
high number of deaths, yellow denotes medium number of deaths and lastly
green denotes low number of deaths. From the graph, we can see that
countries like USA, Russia, India and parts of Canada, the number of
casualties due to aircraft accidents is very high. On the contrary, in
some countries of Africa and parts of South America, the number of
deaths is on the lower side.

This could be explained by the fact that countries like USA, Russia and
India are richer than those of Africa and parts of South America and
thus have a bigger aircraft network. Flights in those countries are more
frequent than the poorer countries and thus the number of deaths is
distributed in such a way.

Unfortunately we could not obtain any data on the air traffic in
different countries of the world to support our hypothesis. Hence we can
guess at the best the reason for such a distribution of deaths in our
plot.

\newpage

Now to see the **trend of deaths over time** we have made the following 
scatter plot. The graph shows number of deaths every year from 1938-2022.

```{r echo=FALSE}
data <- final_df %>% group_by(Date) %>% summarise(sum(`Total deaths`), 
                                                  sum(`Crew deaths`),
                                                  sum(`Passenger deaths`),
                                                  sum(`Ground deaths`))
colnames(data) <- c("Date", "Total deaths","Crew deaths", "Passenger deaths", "Ground deaths")

deaths_time <- ggplot(data, aes(x = Date, y = `Total deaths`)) +
  geom_point(col = "red") +
  geom_line() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Deaths vs Year", x = "Year", y = "Total deaths") +
  theme_light()
deaths_time

```


It can be observed that there is an **overall increase** in the number of deaths
over time. But if we divide the time into smaller periods, a **gradual increase** can be
observed from **1938-1970**. Then, there is an **extreme observation** around the year
**2000**. After that, a **decreasing trend** in the number of deaths can be observed.
This might be due to advancement in technology which has made
airplanes safer, thus reducing the number of deaths. 

\newpage

To observe which company was the safest to travel over the years, we made the following
bar plot. Here the height of the bars represent the **average number of deaths per flight**
for a particular company.

```{r echo=FALSE, fig.width=11}
temp <- final_df
vec <- numeric(length = 544)
vec <- sapply(vec, function(x) 1)
temp$freq <- c(vec)

data_company <- temp %>% group_by(Aircraft) %>% summarise(sum(`Total deaths`), 
                                                       sum(`Crew deaths`),
                                                       sum(`Passenger deaths`),
                                                       sum(`Ground deaths`),
                                                       sum(`freq`))
colnames(data_company) <- c("Aircraft", "Total deaths","Crew deaths", "Passenger deaths", "Ground deaths", "freq")
for(i in 2:5)
{
  data_company[ ,i] = data_company[ ,i]/data_company$freq
}

death_company <- ggplot(data_company, aes(x = Aircraft, y = `Total deaths`, fill = Aircraft)) + 
  geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Number of deaths vs Aircraft companies") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(y = "Average number of deaths")
death_company

```


From the graph, we can easily observe that **Airbus** has the highest average deaths per flight, immediately followed by **Boeing**. **Latécoère** and **Bombardier** lie on the other
end of the spectrum with the least number of average deaths per flight.

We can also make the plot of the **total number of deaths caused for each company**, but that will give us **wrong analysis**. The reason is simple, maybe a company has very large number of flights compared to others, hence resulting in more number of accidents. Thus studying the total number of deaths against various manufacturers would give us a biased result and would spread a false information.

\newpage

To answer our **final question**, we made scatter plots of total deaths over the years for each company. Here are few plots for some of the major companies such as Airbus, Boeing, Tupolev etc. 

```{r echo=FALSE, fig.height=3, fig.width=3}

# Airbus---------------------

type_1 <- final_df$`Total deaths`[which(final_df$Aircraft == "Airbus")]
type_2 <- final_df$Aircraft[which(final_df$Aircraft == "Airbus")]
df_types <- data.frame(type_1, type_2, final_df$Date[which(final_df$Aircraft == "Airbus")])

ggplot(df_types, aes(df_types[,3], df_types[,1]))+
  geom_point()+
  geom_smooth(method = 'lm',
              formula = y~x,
              se = F)+
  labs(x = "Year", y = "Number of Deaths", title = "Airbus") +
  theme_minimal() +
  ylim(0,300)

# Antonov-------------------------------------------------------------

type_1 <- final_df$`Total deaths`[which(final_df$Aircraft == "Antonov")]
type_2 <- final_df$Aircraft[which(final_df$Aircraft == "Antonov")]
df_types <- data.frame(type_1, type_2, final_df$Date[which(final_df$Aircraft == "Antonov")])

ggplot(df_types, aes(df_types[,3], df_types[,1]))+
  geom_point()+
  geom_smooth(method = 'lm',
              formula = y~x,
              se = F)+
  labs(x = "Year", y = "Number of Deaths", title = "Antonov") +
  theme_minimal() +
  ylim(0,300)

# Boeing------------------------------------

type_1 <- final_df$`Total deaths`[which(final_df$Aircraft == "Boeing")]
type_2 <- final_df$Aircraft[which(final_df$Aircraft == "Boeing")]
df_types <- data.frame(type_1, type_2, final_df$Date[which(final_df$Aircraft == "Boeing")])

ggplot(df_types, aes(df_types[,3], df_types[,1]))+
  geom_point()+
  geom_smooth(method = 'lm',
              formula = y~x,
              se = F)+
  labs(x = "Year", y = "Number of Deaths", title = "Boeing") +
  theme_minimal() +
  ylim(0,300)

# Douglas--------------------------------------

type_1 <- final_df$`Total deaths`[which(final_df$Aircraft == "Douglas")]
type_2 <- final_df$Aircraft[which(final_df$Aircraft == "Douglas")]
df_types <- data.frame(type_1, type_2, final_df$Date[which(final_df$Aircraft == "Douglas")])

ggplot(df_types, aes(df_types[,3], df_types[,1]))+
  geom_point()+
  geom_smooth(method = 'lm',
              formula = y~x,
              se = F)+
  labs(x = "Year", y = "Number of Deaths", title = "Douglas") +
  theme_minimal() +
  ylim(0,300)


# Ilyushin-------------------------

type_1 <- final_df$`Total deaths`[which(final_df$Aircraft == "Ilyushin")]
type_2 <- final_df$Aircraft[which(final_df$Aircraft == "Ilyushin")]
df_types <- data.frame(type_1, type_2, final_df$Date[which(final_df$Aircraft == "Ilyushin")])

ggplot(df_types, aes(df_types[,3], df_types[,1]))+
  geom_point()+
  geom_smooth(method = 'lm',
              formula = y~x,
              se = F)+
  labs(x = "Year", y = "Number of Deaths", title = "Ilyushin") +
  theme_minimal() +
  ylim(0,300)

# Lockheed-------------------------------

type_1 <- final_df$`Total deaths`[which(final_df$Aircraft == "Lockheed")]
type_2 <- final_df$Aircraft[which(final_df$Aircraft == "Lockheed")]
df_types <- data.frame(type_1, type_2, final_df$Date[which(final_df$Aircraft == "Lockheed")])

ggplot(df_types, aes(df_types[,3], df_types[,1]))+
  geom_point()+
  geom_smooth(method = 'lm',
              formula = y~x,
              se = F)+
  labs(x = "Year", y = "Number of Deaths", title = "Lockheed") +
  theme_minimal() +
  ylim(0,300)

# McDonnell Douglas-----------------------

type_1 <- final_df$`Total deaths`[which(final_df$Aircraft == "McDonnell Douglas")]
type_2 <- final_df$Aircraft[which(final_df$Aircraft == "McDonnell Douglas")]
df_types <- data.frame(type_1, type_2, final_df$Date[which(final_df$Aircraft == "McDonnell Douglas")])

ggplot(df_types, aes(df_types[,3], df_types[,1]))+
  geom_point()+
  geom_smooth(method = 'lm',
              formula = y~x,
              se = F)+
  labs(x = "Year", y = "Number of Deaths", title = "McDonnell Douglas") +
  theme_minimal() +
  ylim(0,300)

# Tupolev------------------------------

type_1 <- final_df$`Total deaths`[which(final_df$Aircraft == "Tupolev")]
type_2 <- final_df$Aircraft[which(final_df$Aircraft == "Tupolev")]
df_types <- data.frame(type_1, type_2, final_df$Date[which(final_df$Aircraft == "Tupolev")])

ggplot(df_types, aes(df_types[,3], df_types[,1]))+
  geom_point()+
  geom_smooth(method = 'lm',
              formula = y~x,
              se = F)+
  labs(x = "Year", y = "Number of Deaths", title = "Tupolev") +
  theme_minimal() +
  ylim(0,300)


```



From the above graphs following oberservations can be made:

1.  Here we can see that the number of deaths for Airbus flights have
    decreased over the period of time.
2.  The number of deaths for Antonov has remained more or less unchanged with time.
3.  **Boeing** and **Douglas** have recorded increasing number of deaths over the time period and so did the     other companies.

So we can say that maybe **Airbus** has made more improvements in their
air crafts as compared to others over time.

\newpage
```{=tex}
\begin{center} \Large \textbf{Conclusions}\end{center}
```
By analyzing our data set we have discovered some compelling patterns for airplane accidents. The results can be summarized in the following points:

1.    Most of the deaths have occurred for **Commercial** aircrafts followed by **Military** aircrafts.
2.   Accidents  during the **En route** phase have resulted in the most number of deaths.
3.    The number of deaths due to aircraft accidents have been on a **decrease** during the last few years.
4.    **Airbus** and **Boeing** have the highest number of average deaths during an accident.
5.    However, **Airbus** has made some quality improvements in their planes overtime.

The main objective of this project was to raise awareness of flight safety and understand the progress of the aviation industry. The aircraft industry has come a long way but it still has a lot to improve.

\newpage

```{=tex}
\begin{center} \Large \textbf{References}\end{center}
```
1.    The data set was obtained from Wikipedia [List of deadliest aircraft accidents and incidents]
\newline
<https://en.wikipedia.org/wiki/List_of_deadliest_aircraft_accidents_and_incidents>.
2.    The reference for plots was taken from r-statistics [Top 50 ggplot2 Visualizations - The Master List]
\newline
<http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>.














